---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "tools/README-"
)
```
[![CRAN status](https://www.r-pkg.org/badges/version/noaaoceans)](https://cran.r-project.org/package=noaaoceans)
[![Build Status](https://travis-ci.org/warlicks/NOAATides.svg?branch=master)](https://travis-ci.org/warlicks/NOAATides)
[![codecov](https://codecov.io/gh/warlicks/NOAATides/branch/master/graph/badge.svg)](https://codecov.io/gh/warlicks/NOAATides)

  
# noaaoceans

The **noaaoceans** package is designed to facilitate access to various National Oceanic and Atmospheric Administration (NOAA) data sources.  The current version focuses on accessing data from the [CO-OPS API](https://tidesandcurrents.noaa.gov/api/).  The package also facilitates the collection of basic metadata for each of the stations that collect the data available in the API.  

## Example

### Find all the CO-OPS stations in Washington. 
```{r find_station, message=FALSE, warning=FALSE}
library(noaaoceans)
library(dplyr)
library(ggplot2)
library(maps)
library(mapdata)

# Get a list of all the stations.
station_df <- list_coops_stations()

# Filter to stations in Washington with Water Temp Sensor
wa_station <- station_df %>% 
    filter(station_state == 'WA' & water_temp == '1')
wa_station %>% dplyr::as_tibble(.) %>% head()
```

### Query API For Water Temperature Data
```{r query_api}
# Create an empty storage data frame. 
water_temp <- data.frame()

# Loop through the station and call the API for each station
for (i in wa_station$station_id) {
    query_df <- query_coops_data(station_id = i,
                                 start_date = '20181201',
                                 end_date = '20181231',
                                 data_product = 'water_temperature',
                                 interval = 'h') # hourly readings
    
    # Add current station results to the storage data frame 
    water_temp <- water_temp %>% bind_rows(., query_df)
}

water_temp %>% as_tibble(.) %>% head()
```

### Plot the Average Water Temp
```{r plot_data, fig.height=5, fig.width=7}
# Correct data types. 
water_temp <- water_temp %>% 
    mutate(v = as.numeric(v), t = as.POSIXct(t))

water_temp %>% 
    group_by(t) %>% 
    # Compute the hourly average. 
    summarise(avg_temp = mean(v, na.rm = TRUE)) %>% 
    # Plot the hourly average. 
    ggplot(aes(x = t, y = avg_temp)) +
    geom_path() +
    labs(x = "Date",
         y = 'Average Water Temperature',
         title = 'Average Hourly Water Temperature In Washington During December 2018')
```
